---
title: "PCA"
author: "Pumla"
date: "2024-06-26"
output: html_document
---

#Setting up environment 
#Clean environment
```{r}
rm(list = ls(all.names = TRUE)) 
gc() 
options(max.print = .Machine$integer.max, scipen = 999, stringsAsFactors = F, dplyr.summarise.inform = F)
```


#Load libraries
```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(pheatmap)
```

#Set the working directory
```{r}
setwd("~/starmm/pca")
```


```{r}
fpkm_data <- read.csv("FPKM.csv", header = TRUE, stringsAsFactors = FALSE)
```


#Read design.csv containing SampleID and Treatment
```{r}
design_data <- read.csv("Designpca.csv", header = TRUE, stringsAsFactors = FALSE)
```


#View the contents of fpkm_data
```{r}
View(fpkm_data)
```


#View the contents of design_data
```{r}
View(design_data)
```


#Display the structure of fpkm_data
```{r}
str(fpkm_data)
```


#Display the first few rows of fpkm_data
```{r}
head(fpkm_data)
```


#Display the structure of design_data
```{r}
str(design_data)
```


#Extract numeric data for PCA (assuming columns 2 to end are numeric)
```{r}
fpkm_numeric <- fpkm_data[, -1]
```


#Log2 transformation
```{r}
fpkm_log <- log2(fpkm_numeric + 1)
```


#Transpose data for PCA (assuming rows are samples and columns are features)
```{r}
fpkm_log_t <- t(fpkm_log)
```


#Perform PCA
```{r}
pca_result <- prcomp(fpkm_log_t, center = TRUE, scale. = FALSE)
```


#Extract PCA scores
```{r}
pc_scores <- as.data.frame(pca_result$x)
```


#Add Treatment information to PCA scores based on SampleID
```{r}
pc_scores$SampleID <- rownames(pc_scores)
pc_scores_with_treatment <- merge(pc_scores, design_data, by = "SampleID")
```


#Create PCA plot
```{r}
p <- ggplot(pc_scores_with_treatment, aes(x = PC1, y = PC2, color = Treatment, label = SampleID)) +
  geom_point(size = 2) +
  geom_text(size = 3, hjust = -0.2, vjust = 0, color = "black") +
  scale_color_manual(values = c("healthy" = "#1f78b4", "treated" = "#b2df8a")) +
  labs(x = paste("PC1 (", round(summary(pca_result)$importance[2, 1] * 100, 2), "%)"),
       y = paste("PC2 (", round(summary(pca_result)$importance[2, 2] * 100, 2), "%)"),
       color = "Treatment") +
  theme_minimal()
```


#Display the plot
```{r}
print(p)
```


#to remove print 
```{r}
ggplot(pc_scores_with_treatment, aes(x = PC1, y = PC2, color = Treatment, label = SampleID)) +
  geom_point(size = 4) +
  geom_text(size = 3, hjust = -0.2, vjust = 0, color = "black") +
  scale_color_manual(values = c("healthy" = "#1f78b4", "treated" = "#b2df8a")) +
  labs(x = paste("PC1 (", round(summary(pca_result)$importance[2, 1] * 100, 2), "%)"),
       y = paste("PC2 (", round(summary(pca_result)$importance[2, 2] * 100, 2), "%)"),
       color = "Treatment") +
  theme_minimal()
```







